A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution \thanks{A preliminary version of this work has appeared in the Proceedings of the 26th International Symposium on Graph Drawing and Network Visualization (GD 2018). This work is supported by the DFG grant Ka812/17-1}

### 2018-09-16 paper/journal.tex
Working a bit more on the journal version

% ==================================================================
\subsection{Avoiding local maxima}
\label{ssec:bottlenecks}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behaviour. What would one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{itemize}
\item[--] if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} by including all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item[--] afterwards, the vertex-pool switches back to the critical vertices.
\end{itemize}

@@

While our unrestricted algorithm produces drawings with better crossing resolution, this comes at a cost of drastically increased aspect ratio (see Fig.~\ref{fig:cr-res-2}), which, however, is still better that the corresponding aspect ratio of the drawings produced by the algorithm of Argyriou et al. For the latter algorithm, it seems that the forces due to the angles formed at the crossings outperform the corresponding spring forces, which try to keep the lengths of the edges short. Going back to our unrestricted algorithm, its behaviour is up to a certain degree expected, mainly due to the fact that there is no control on the lengths of the edges (as it is, e.g., in the case of the force-directed algorithm of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, whose sping forces inpose stong restrictions on the edge lengths). On the other hand, the restricted variant of our algorithm, which does not allow the aspect ratio to increase, has more or less comparable performance (in terms of aspect ratio) as the one of Huang et al.

@@

A different behaviour can be observed in the number of iterations, which are required by the algorithms to converge; refer to Fig.~\ref{fig:cr-res-4}. We note here that we used different criteria to determine whether the algorithms of our experiment had converged. For our algorithms and for the force-directed algorithm by Huang et al., we assumed that the algorithm had converged, if the crossing resolution between 500 consecutive iterations was not improved by more than 0.001 degrees. For the algorithm by Argyriou et al., we decided to use a much more restricted convergence criterion, because the produced layouts can change vastly between  consecutive iterations. We made this choice mainly to have ``comparable'' number of iterations among the algorithms of the experiment. In this direction, we adopted the convergence criterion that the authors used in their previous experimental analysis, that is, we assumed that the algorithm had converged, if the crossing resolution between two consecutive iterations was not improved by more than 0.001 degrees. Observe that even under this different convergence criterion, the algorithm needs significantly more iterations to converge than the remaining three algorithms of the experiment; see Fig.~\ref{fig:cr-res-4}.

@@

\myparagraph{Crossing resolution} As with the Rome graphs, we observed again that both variants of our algorithm again outperformed the two force-directed algorithms; see Fig.~\ref{fig:northCrossing-1}. Remarkable is the synchronous behaviour of all four algorithms regarding the crossing resolution, as the curves are nearly parallel. The results actually indicate that one can classify these graphs into ``hard'' and ``easy'' ones, when maximizing their crossing resolution. In particular, the graphs with 50 to 70 vertices appear to be harder than the graphs with 70 to 80 vertices. As expected, the aspect ratio of the drawings produced by both the unrestricted and the restricted variants of our algorithm depends vastly on the number of vertices of the graphs. However, this natural behaviour cannot be observed in the resulting drawings of the two force-directed algorithms, which appear to be quite unstable resulting in a large variance. Again the restricted variant of our algorithm and the two force directed approaches produce drawings with similar aspect ratio, which is much lower than the one of our unrestricted algorithm for larger graphs. All four algorithms behave nearly the same in terms of the number of crossings; see Fig.~\ref{fig:northCrossing-3}. In terms of the number of iterations, we observe that somewhat surprisingly the algorithm of Argyriou et al. converges using the least amount of iterations, while the remaining three algorithms behave nearly the same; see Fig.~\ref{fig:northCrossing-4}.

\myparagraph{Total resolution} For small graphs, we observed similar results as in the experiment on the Rome graphs, that is, our unrestricted algorithm outperforms (in terms of the total resolution) the other three algorithms of the experiment, while the restricted variant of our algorithm is slightly better than the algorithms by Argyriou et al.\ and by Huang et al.; see Fig.~\ref{fig:northTotal-1}. For larger graphs, however, the two variants of our algorithm achieve similar results while still outperforming the algorithm by Huang et al. Regarding the aspect ratio and the number of crossings of the produced drawings, we observe that all algorithms have more or less the same behaviour as when adjusted

The results for the aspect ratio and number of crossings are similar to those observed  the crossing resolution experiment, with the exception of the fact that the algorithm of Huang et al. performs more stable with respect to the aspect ratio; see Figs.~\ref{fig:northTotal-2} and~\ref{fig:northTotal-3}. With respect to the number of iterations, our two algorithms and the one by Argyriou et al. show similar behavior needing more iterations than the algorithm by Huang et al.\ in order to converge; see Fig.~\ref{fig:northTotal-4}. %Observe that the number of iterations does not seem to correlate with the number of vertices.

In the angular resolution experiment, we again obtain a not-so-clear picture concerning the ranking of the algorithms, especially for higher number of vertices the ranking varies; see Fig.~\ref{fig:northAngular-1}. Only the algorithm by Huang et al. seems to be mostly at the last rank. Concerning the aspect ratio we see very good behaviour for our restricted variant and the algorithm by Huang et al. while the remaining two algorithms show large variance and much worse values; see Fig.~\ref{fig:northAngular-2}. For the number of crossings, we again observe that all algorithms achieve similar values, however, our unrestricted algorithm and the algorithm by Argyriou et al. achieve slightly higher values for larger graphs; see Fig.~\ref{fig:northAngular-3}. Finally, both our algorithms and the one by Huang et al. need a similar number of iterations for convergence which is lower than the one by Argyriou et al.; see Fig.~\ref{fig:northAngular-4}.

@@

In addition to the experiments described in Section~\ref{sec:experiments}, we also evaluated how our algorithm performs, if we restrict its vertices to integer grid coordinates. In particular, we were interesting to see how the different quality measures that we evaluated in Section~\ref{sec:experiments} are affected by the restrictions imposed on having the vertices of the graph on integer grids of different sizes:
%
\begin{inparaenum}[(i)]
\item $10^6 \times 10^6$
\item $10^4 \times 10^4$
\item $10^3 \times 10^3$, and
\item $10^2 \times 10^2$.
\end{inparaenum}
%
The test suite for this experiment was again the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}. However, since our algorithm is guaranteed to produce a grid drawing, only if its initial drawing is grid, each of the Rome graphs was initially laid out by randomly placing its vertices on the grid, ensuring that neither two vertices nor two edges overlap. The layouts for each of the different sizes of the grid were computed with the variant of our algorithm that optimizes the crossing resolution; Fig.~\ref{fig:experimentsGrid} summarizes the results.

Regarding the crossing resolution, we can observe that with increasing grid size, we could achieve better crossing resolution; see Fig.~\ref{fig:crossingResolutionGrid}. More precisely, a grid of size $10^2 \times 10^2$ was too restrictive for the vast majority of the graphs. As a result, the reported drawings were often the initial ones (as our algorithm could not improve them), especially for large graphs. Significantly fewer were the graphs for which our algorithm could not report an improved drawing, when the grid size was set to $10^3 \times 10^3$.
%Starting from grid size $10^3 \times 10^3$, we could observe, for small graphs, the performance was close to the performance of the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}).
%For grid size $10^3 \times 10^3$, the performance declined for the larger graphs in the test set and we again could observe that for some of them the algorithm was not able to improve the initial layout.
For grid size $10^4 \times 10^4$, the drawings produced by our algorithm were on average by only $10^\circ$ worse than those produced by the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}), while the gap was closer for grid size $10^6 \times 10^6$.

@@

As expected, the smaller the underlying grid is, the more crossings the computed drawings contain; see Fig.~\ref{fig:crossingResolutionCrossingNumberGrid}. As a result, the unrestricted variant of our algorithm clearly outperforms all other ones. It is worth noting that the differences are clear between grid sizes $10^2 \times 10^2$, $10^3 \times 10^3$ and $10^4 \times 10^4$. Notably, there is only a slight improvement (in terms of the number of crossings) from grid size $10^4 \times 10^4$ to $10^6 \times 10^6$. %This can partially be explained with the different choice of the initial drawing.
%
On the other hand, the number of iterations needed for convergence increases with the grid size (see Fig.~\ref{fig:crossingResolutionTimeGrid}), with the exception of the grid of size $10^2 \times 10^2$, which verifies our previous observation that for the vast majority of the graphs an improved drawing could not be reported.

In conclusion, we can state that our algorithm is still able to compute drawings with high crossing resolution when restricted to a grid, as long as the grid is not too small. However, the computation of a grid drawing takes longer, which is of course expected.
%than in the unrestricted version (up to five times as long for grid size $10^4 \times 10^4$ or eight times as long for grid size $10^6 \times 10^6$). In order
Finally, note that the choice of the initial grid drawing seems to affect the performance of our algorithm, both with respect to the quality of the produced drawings (counted here in terms of the crossing resolution) but also with respect to the number of iterations needed to converge.
%To improve the performance of our algorithm in this restricted variant, it may be interesting to see how much the choice of the initial drawing affects the resulting crossing resolution. Further experiments in this direction are needed.

>>>>

% ==================================================================
\subsection{Avoiding local maxima}
\label{ssec:bottlenecks}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behavior. What would one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{itemize}
\item[--] if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} by including all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item[--] afterwards, the vertex-pool switches back to the critical vertices.
\end{itemize}

@@

While our unrestricted algorithm produces drawings with better crossing resolution, this comes at a cost of drastically increased aspect ratio (see Fig.~\ref{fig:cr-res-2}), which, however, is still better that the corresponding aspect ratio of the drawings produced by the algorithm of Argyriou et al. For the latter algorithm, it seems that the forces due to the angles formed at the crossings outperform the corresponding spring forces, which try to keep the lengths of the edges short. Going back to our unrestricted algorithm, its behavior is up to a certain degree expected, mainly due to the fact that there is no control on the lengths of the edges (as it is, e.g., in the case of the force-directed algorithm of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, whose sping forces inpose stong restrictions on the edge lengths). On the other hand, the restricted variant of our algorithm, which does not allow the aspect ratio to increase, has more or less comparable performance (in terms of aspect ratio) as the one of Huang et al.

@@

A different behavior can be observed in the number of iterations, which are required by the algorithms to converge; refer to Fig.~\ref{fig:cr-res-4}. We note here that we used different criteria to determine whether the algorithms of our experiment had converged. For our algorithms and for the force-directed algorithm by Huang et al., we assumed that the algorithm had converged, if the crossing resolution between 500 consecutive iterations was not improved by more than 0.001 degrees. For the algorithm by Argyriou et al., we decided to use a much more restricted convergence criterion, because the produced layouts can change vastly between  consecutive iterations. We made this choice mainly to have ``comparable'' number of iterations among the algorithms of the experiment. In this direction, we adopted the convergence criterion that the authors used in their previous experimental analysis, that is, we assumed that the algorithm had converged, if the crossing resolution between two consecutive iterations was not improved by more than 0.001 degrees. Observe that even under this different convergence criterion, the algorithm needs significantly more iterations to converge than the remaining three algorithms of the experiment; see Fig.~\ref{fig:cr-res-4}.

@@

\myparagraph{Crossing resolution} As with the Rome graphs, we observed again that both variants of our algorithm again outperformed the two force-directed algorithms; see Fig.~\ref{fig:northCrossing-1}. Remarkable is the synchronous behavior of all four algorithms regarding the crossing resolution, as the curves are nearly parallel. The results actually indicate that one can classify these graphs into ``hard'' and ``easy'' ones, when maximizing their crossing resolution. In particular, the graphs with 50 to 70 vertices appear to be harder than the graphs with 70 to 80 vertices. As expected, the aspect ratio of the drawings produced by both the unrestricted and the restricted variants of our algorithm depends vastly on the number of vertices of the graphs. However, this natural behavior cannot be observed in the resulting drawings of the two force-directed algorithms, which appear to be quite unstable resulting in a large variance. Again the restricted variant of our algorithm and the two force-directed algorithms produce drawings with similar aspect ratio, which is much lower than the one of our unrestricted algorithm for larger graphs. All four algorithms behave nearly the same in terms of the number of crossings; see Fig.~\ref{fig:northCrossing-3}. In terms of the number of iterations need for convergence, we observe that somewhat surprisingly the algorithm of Argyriou et al.\ converges using the least amount of iterations, while the remaining three algorithms behave nearly the same; see Fig.~\ref{fig:northCrossing-4}.

\myparagraph{Total resolution} For small graphs, we observed similar results as in the experiment on the Rome graphs, that is, our unrestricted algorithm outperforms (in terms of the total resolution) the other three algorithms of the experiment, while the restricted variant of our algorithm is slightly better than the algorithms by Argyriou et al.\ and by Huang et al.; see Fig.~\ref{fig:northTotal-1}. For larger graphs, however, the two variants of our algorithm achieve similar results, while still outperforming the algorithm by Huang et al. Regarding the aspect ratio and the number of crossings of the produced drawings, we observe that all algorithms have more or less the same behavior as when adjusted for maximizing the crossing resolution only. The only exception can be observed in the algorithm by Huang et al., which performs more stable; see Figs.~\ref{fig:northTotal-2} and~\ref{fig:northTotal-3}. With respect to the number of iterations required for convergence, the two variants of our algorithms and the force-directed algorithm by Argyriou et al.\ show similar behavior. The algorithm by Huang et al., on the other hand, needs less iterations for convergence; see Fig.~\ref{fig:northTotal-4}. Note, however, that there is no immediate correlation with the number of vertices of the graph.

\myparagraph{Angular resolution} Here, the picture is not so clear concerning the ranking of the algorithms; see Fig.~\ref{fig:northAngular-1}. Only the algorithm by Huang et al.\ seems to be mostly at last in the rank. Concerning the aspect ratio, we see very good behavior for the restricted variant of our algorithm and the algorithm by Huang et al., while the remaining two algorithms show large variance and much worse values; see Fig.~\ref{fig:northAngular-2}. Regarding the number of crossings, we again observe that all algorithms have similar behavior. Note, however, that the unrestricted variant of our algorithm and the algorithm by Argyriou et al.\ yield drawings with more crossing, especially for graphs with large number of vertices; see Fig.~\ref{fig:northAngular-3}. Finally, both variants of our algorithm and the force-directed algorithm by Huang et al.\ need comparable number of iterations for convergence, which most of the times is lower than the corresponding one by Argyriou et al.; see Fig.~\ref{fig:northAngular-4}.

@@

In addition to the experiments that we have described so far, we also evaluated how our algorithm performs, if we restrict its vertices to integer grid coordinates. In particular, we were interesting to see how the different quality measures that we have evaluated so far are affected by the restrictions imposed on having the vertices of the graph on integer grids of different sizes:

\begin{enumerate}[(i)]
\item $10^6 \times 10^6$
\item $10^4 \times 10^4$
\item $10^3 \times 10^3$, and
\item $10^2 \times 10^2$.
\end{enumerate}

The test suite for this experiment was the non-planar Rome graphs. The algorithm to be evaluated was the variant of our algorithm that optimizes the drawing's crossing resolution. However, since this algorithm is guaranteed to produce a grid drawing, only if its initial drawing is grid, each of the Rome graphs was initially laid out by randomly placing its vertices on the grid, ensuring that neither two vertices nor two edges overlap. Fig.~\ref{fig:experimentsGrid} summarizes the results.

Regarding the crossing resolution, we natural observe that the bigger the grid is, the easier it is to compute a drawing with higher crossing resolution; see Fig.~\ref{fig:crossingResolutionGrid}. More precisely, the grid of size $10^2 \times 10^2$ was too restrictive for the vast majority of the graphs. As a result, the reported drawings were often the initial ones (as our algorithm could not improve them), especially for large graphs. Significantly fewer were the graphs for which our algorithm could not report an improved drawing, when the underlying grid was of size $10^3 \times 10^3$. The crossing resolution of the drawings produced by our algorithm, when the underlying grid was of size $10^4 \times 10^4$, were on average worse by only $10^\circ$ than those produced by our algorithm, when imposing no restriction on the size of the underlying grid (refer to the double line in Fig.~\ref{fig:crossingResolutionGrid}). As expected, the gap was more narrow, when the underlying grid was of size $10^6 \times 10^6$.

@@

As expected, the smaller the underlying grid is, the more crossings the computed drawings contain; see Fig.~\ref{fig:crossingResolutionCrossingNumberGrid}. As a result, the drawings produced by our algorithm, when imposing no restriction on the size of the underlying grid, are clearly better (in terms of the number of crossings) than those produced by our algorithm, when the size of the underlying grid is limited. As a matter of fact, the larger the underlying grid is, the fewer crossing the reported drawings have, and the differences are quite clear. Exceptional are the grids of sizes $10^4 \times 10^4$ and $10^6 \times 10^6$, where the improvement (in the number of crossings) from the latter to the former is small on average; this can partially be explained by the different choices of the initial drawings.

In general, the number of iterations needed for convergence increases with the size of the underlying grid; see Fig.~\ref{fig:crossingResolutionTimeGrid}. Exceptional seems to be the $10^2 \times 10^2$ grid, which verifies our previous observation that for the vast majority of the graphs an improved drawing could not be reported.

In conclusion, we can state that our algorithm is still able to compute drawings with high crossing resolution, when restricted to a grid, as long as the grid is not too small. However, the computation of a grid drawing takes longer, which is of course expected. Finally, the choice of the initial grid drawing seems to affect the performance of our algorithm, both with respect to the quality of the produced drawings (counted here in terms of the crossing resolution), but also with respect to the number of iterations needed for convergence.
